---
  title: "Clean"
author: "Ewout Knoester & Gian Lu Dado"
date: "11 April 2023"
output: html_document
---
  
  # Setup
  ```{r setup, include=FALSE}
rm(list=ls()) # Clear workspace
knitr::opts_knit$set(root.dir = '/tmp') #Set directory at current directory for all subsequent chunks

setwd("~/Documents/Analysis/PDAMICORNIS")
getwd()

library(readxl) # Import excel sheets
library(tidyverse) # Tidy data
library(panelr) # Convert data from wide to long
library(nlme) # GLS
library(emmeans) # Pairwise comparisons
library(betareg) # Beta regression
library(ggthemes) # Pretty plots
library(car) # ANOVA results GLM
library(NCmisc) # Check packages used
library(writexl)
library(cowplot) # Combine plots
library(e1071) # Determine skewness coefficient

# Function to facilitate averaging datasets
data_summary <- function(data, varname, groupnames){
  require(plyr)
  summary_func <- function(x, col){
    c(mean = mean(x[[col]], na.rm=TRUE),
      sd = sd(x[[col]], na.rm=TRUE),
      n  = length(x[[col]]),
      se = sd(x[[col]], na.rm=TRUE)/sqrt(length(x[[col]])))
  }
  data_sum<-ddply(data, groupnames, .fun=summary_func,
                  varname)
  data_sum <- rename(data_sum, c("mean" = varname))
  return(data_sum)
}
```

# Data selection (preparing dataset for archiving & sharing)
```{r data selection, include=FALSE}

# Import excel sheet
raw.data <- read_excel("Coral growth - experiment Pdamicornis 2022.xlsx", sheet = "DATA") 

# Insert new columns to calculate average SGR over a whole period while maintaining same wide-format style data after the last column of raw.data
raw.data.1 <- as.data.frame(append(raw.data, list(Date_3 = "2023-03-22", Length_3 = NA, Width1_3 = NA, Width2_3 = NA, EV_3 = NA, Condition_3 = NA, Cause_3 = NA,SGR_3 = NA, Comments_3 = NA), after = (ncol(raw.data))))

## Calculate SGR (Specific Growth Rate) over whole period and per month
raw.data.1$SGR_1<- log(raw.data.1$EV_1/raw.data.1$EV_0)/as.numeric(as.Date(as.character(raw.data.1$Date_1), format="%Y-%m-%d") - as.Date(as.character(raw.data.1$Date_0), format="%Y-%m-%d"))

raw.data.1$SGR_2 <-log(raw.data.1$EV_2/raw.data.1$EV_1)/as.numeric(as.Date(as.character(raw.data.1$Date_2), format="%Y-%m-%d") - as.Date(as.character(raw.data.1$Date_1), format="%Y-%m-%d"))

raw.data.1$SGR_3<- log(raw.data.1$EV_1/raw.data.1$EV_0)/as.numeric(as.Date(as.character(raw.data.1$Date_1), format="%Y-%m-%d") - as.Date(as.character(raw.data.1$Date_0), format="%Y-%m-%d"))

## Fill in end Condition for Average time factor (NB: end Condition will be used in analysis)
raw.data.1$Condition_3 <- raw.data.1$Condition_1
raw.data.1$Cause_3 <- raw.data.1$Cause_1

# Convert to long data frame
raw.data.2 <- as.data.frame(long_panel(raw.data.1, prefix = "_" , begin = 0, end = 3, label_location = "end"))

raw.data.2$Cause[raw.data.2$Condition > 0] <- NA # All fragments with Condition>0.0 are considered alive

## Set species and rename
species.order = c("Pocillopora damicornis")
  
raw.data.2 %<>% 
  mutate(Species = Species %>% 
           factor(levels = species.order))

## Read in NAs properly
raw.data.2$SGR[raw.data.2$SGR == "NA"] <- NA
raw.data.2$Cause[raw.data.2$Cause == "NA"] <- NA

## Set correct data types
raw.data.2$SGR <- as.numeric(raw.data.2$SGR)
raw.data.2$Date <- format(raw.data.2$Date, "%m/%Y")

## Merge overlapping months for visualization purposes
date.order = c("11/2022", "12/2022","02/2023","03/2023") 

raw.data.2 %<>%
  mutate(Date = Date %<>% 
           factor(levels = date.order)%>% 
  fct_recode('Average' = '03/2023'))

# Drop and reorder some columns
clean.data <- raw.data.2 %>%
  dplyr::select(c("id", "Structure", "Position", "Species", 
           "Date", "Length", "Cause", "EV","Condition", "SGR")) 

# Tidy formatting data frame
clean.data$Length <- round(clean.data$Length, 1)
clean.data$EV <- round(clean.data$EV, 0)
clean.data$SGR <- round(clean.data$SGR, 3)

# Remove data from February 2023
clean.data <- subset(clean.data, Date != "02/2023")

# Export data selection
write_xlsx(clean.data, "PD_Cleaning experiment.xlsx") # Save clean.data as a new excel document.

```

# Data cleaning
```{r data cleaning}

# Import excel sheet
clean.data.1 <- read_excel("PD_Cleaning experiment.xlsx", sheet = 1)

# Set correct data types
clean.data.1$Structure <- factor(clean.data.1$Structure)
clean.data.1$Species <- as.factor(clean.data.1$Species)
clean.data.1$SGR <- as.numeric(clean.data.1$SGR)
clean.data.1$Cause <- factor(clean.data.1$Cause)
clean.data.1$Date <- ordered(clean.data.1$Date, c("11/2022", "12/2022","Average"))
clean.data.1$Position <- as.factor(clean.data.1$Position)

# Create subset to inspect missing values
missing <- subset(clean.data.1, Cause == "missing")
dead<-subset(clean.data.1, Cause == "dead")

## Summary of missing values 
### Indicating that missing fragments are roughly equally distributed across Treatments
summary(missing)
summary(dead)

#! Decision to leave out missing fragments from analysis as pattern is not related to Treatment
clean.data.1 <- clean.data.1[-which(clean.data.1$Cause == "missing"),]

write_xlsx(clean.data.1, "PD_Cleaning experiment_new.xlsx") # Save clean.data.1 as a new clean excel document.
```

# SGR analysis
## Model selection
```{r SGR: linear model selection}

clean.data.1 <- read_excel("PD_Cleaning experiment_new.xlsx", sheet = 1)

# DATA PREP
## Create subset for SGR where Condition > 80: only healthy corals will be used to determine SGR
sgr.data <- subset(clean.data.1, Condition >= 80 | is.na(Condition))
sgr.data <- subset(sgr.data, !(is.na(SGR))) # Exclude SGR with NA

# Set correct data types
sgr.data <- sgr.data %>% separate(Structure, c('Treatment', 'Tree'))
sgr.data$Treatment <- factor(sgr.data$Treatment)
sgr.data$Tree<- factor(sgr.data$Tree)
sgr.data$Species<-factor(sgr.data$Species)
sgr.data$Date <- ordered(sgr.data$Date, c("11/2022", "12/2022","Average"))
sgr.data$Position<-as.factor(sgr.data$Position)

## Simplify data by taking average SGR: ignore effect of time (bleaching event/seasons)
sgr.data.1 <- subset(sgr.data, Date != "Average")

## Exclude columns not used
sgr.data.1 <- sgr.data.1 %>%
  dplyr::select(-c("Length", "Cause", "Condition", "EV"))

```

## SGR per date - Model
```{r}

# MODEL
## Random factor to account for non-independence of multiple coral fragments in same nursery Structure
### Full model without random factor
lme.sgr.0r  <- gls(SGR ~ Treatment, method = "REML", na.action = na.omit,
                   data = sgr.data.1) # AIC = -1941.953

### Full model with random factor
lme.sgr.1r  <- lme(SGR ~ Treatment, method = "REML", na.action = na.omit, random = ~1 | Tree, data = sgr.data.1) # AIC = -1942.558

AIC(lme.sgr.0r, lme.sgr.1r) # Tree as random factor improves model substantially and accounts for pseudo-replication

## Allowing for heterogeneity among Treatments
lme.sgr.2r.S  <- lme(SGR ~ Treatment, method = "ML", na.action = na.omit, random = ~1 | Tree, data = sgr.data.1, weights = varIdent(form = ~1 | Species)) #-1989.560

# Allowing for heterogeneity among Treatments
# Model with lmer: fitting more complex mixed models to larger data sets
lmer.sgr.2r.S  <- lmer(SGR ~ Treatment + (1 | Tree), na.action = na.omit, data = sgr.data.1) #-1942.558

AIC(lme.sgr.2r.S, lmer.sgr.2r.S) # Tree as random factor improves model substantially and accounts for pseudo-replication. lme.sgr.2r.S has the lowest AIC, suggesting to be the best model.

## Model output
Anova(lme.sgr.2r.S) # Treatment has a significant effect

```

## SGR per date - Model validation
```{r}
#model validation of linear model of Treatment*Position+Date
#Data is normally distributed
mod <- lme.sgr.2r.S # set model to be validated
op <- par(mfrow = c(2, 4), mar = c(5, 4, 1, 2)) # set layout of figure frame
plot(resid(mod, type = "pearson") ~ fitted(mod)) # fitted vs residuals
abline(0,0)
hist(resid(mod, type = "pearson"), main = "") # histogram residuals
qqnorm(resid(mod, type = "pearson")) # QQplot residuals
plot(sgr.data.1$Treatment, resid(mod, type = "pearson")) # residuals split over Treatment
abline(0,0)
plot(fitted(mod) ~ sgr.data.1$SGR) # response data vs fitted
par(op)
```

## SGR per date - Post hoc and plots
```{r}
``` {r}

# Treatment
# Get averages
sgr.summary.T <- data_summary(sgr.data.1, varname = "SGR", groupnames = c("Treatment"))

# Create ID and order dataframe
sgr.summary.T <- tibble::rownames_to_column(sgr.summary.T, "ID")
sgr.summary.T <- sgr.summary.T[order(as.character(sgr.summary.T$Treatment)),]

# Perform post hoc and get significance letters (Compact Letter Display) between Treatments
hsd.sgr.T <- emmeans(lme.sgr.2r.S, specs = pairwise ~ Treatment, adjust = "tukey")
letters.T <- multcomp::cld(hsd.sgr.T$emmeans, decreasing = TRUE, alpha = 0.05, Letters = letters) # get CLD

# Create ID and order dataframe by name (equals Finder of sgr.summary.SS dataframe)
letters.T <- as.data.frame(append(letters.T,
                                   list(Finder = paste(letters.T$Treatment, letters.T$Species, sep=":")), after = 0))
letters.T <- letters.T[order(as.character(letters.T$Treatment)),]
letters.T <- letters.T %>% mutate(.group = str_squish(.group)) # remove white spaces

# Merge sig.letter dataframe into the summary dataframe
sgr.summary.T <- cbind(sgr.summary.T, siglet = letters.T$.group)
sgr.summary.T <- sgr.summary.T[order(as.numeric(sgr.summary.T$ID)),]

# Plot Treatment bar graph + error bars + letters
SGR_Treatment<-ggplot(sgr.summary.T, aes(x = Treatment, y = SGR))+
  geom_bar(stat = "identity", position = position_dodge())+
 labs(y = expression(paste("SGR (", d^-1,")")), x = "Treatment")+
 scale_y_continuous(limits=c(0, 0.026), breaks = c(0,0.01,0.02,0.03,0.04))+
 geom_errorbar(aes(ymin=SGR-(1*se), ymax=SGR+(1*se)), width=.2, position=position_dodge(.9))+
 geom_text(data = sgr.summary.T, aes(x=Treatment, y = SGR + se*1.5, label = siglet), vjust=0, position=position_dodge(.9))+
 theme_economist()+scale_colour_economist()+
 theme(
   axis.title.x = element_text(color="black", vjust=-2, size = 12, margin = margin(t = 10, r = 0, b = 0, l = 0)),
   axis.text.x=element_text(size=12, face = "bold", vjust=0.5),
   axis.title.y = element_text(color="black" , vjust=4, size = 12, margin = margin(t = 0, r = 10, b = 0, l = 10)),
   axis.text.y=element_text(size=12, face = "bold", vjust=0.5)
   )

SGR_Treatment
ggsave("Pd_SGR_Treatment.pdf") #save image


```


# Survival analysis
## Data preparations
```{r data preparations}

clean.data.1 <- read_excel("PD_Cleaning experiment_new.xlsx", sheet = 1)

# Create survival dataframe
surv.data <- clean.data.1

# Split Structure in two other columns: Treatment, Tree
# Set correct data types
surv.data <- surv.data %>% separate(Structure, c('Treatment', 'Tree'))
surv.data$Treatment <- factor(surv.data$Treatment)
surv.data$Tree<-factor(surv.data$Tree)
surv.data$Species<-factor(surv.data$Species)
surv.data$Date <- ordered(surv.data$Date, c("11/2022", "12/2022","Average"))
surv.data$Position<-as.factor(surv.data$Position)

# Select survival at end of experiment only (ignore effect of time)
surv.data.1 <- subset(surv.data, Date == "Average")

# Exclude columns not used
surv.data.1 <- surv.data.1 %>%
  dplyr::select(-c("Length", "Cause", "SGR", "EV"))

# Get averages per structure, split per Species, Tree and Treatment
surv.avg <- data_summary(surv.data.1, varname = "Condition", groupnames = c("Treatment","Tree"))
surv.avg <- surv.avg %>% 
  dplyr::select(-c("sd", "se"))
surv.avg$Treatment <- factor(surv.avg$Treatment)
surv.avg$Tree<-factor(surv.avg$Tree)

```

## Model selection
  ### Linear model selection
  ```{r AVG Survival}
<!--  
  ### Linear models (lm) couldn't be used: no data transformation could deal with this type of data
  ### Beta regression using full data set didn't work (hopeless residuals), so using data averaged over structures  
  -->
  ### Averaged beta selection
#betareg model selection

# DATA PREP
## Create dataset with average Condition per tree per treatment
surv.avg.2<-surv.avg

hist(surv.avg.2$Condition) # Check if data is normally distributed
skewness(surv.avg.2$Condition) # Skewness coefficient = -2.989741. So, it's negatively (left) skewed

# Inverse transformation on negatively skewed data
surv.avg.2$Condition.t<-sqrt((max(surv.avg.2$Condition+1)-surv.avg.2$Condition))

hist(surv.avg.2$Condition.t)
skewness(surv.avg.2$Condition.t)
# Skewness coefficient = 1.407428

## Transform survival (%) into fraction
surv.avg.2 <- surv.avg.2 %>% mutate(Condition.f = Condition.t/100)

## Re-scale so there are no 0 and 1 in the dataset
surv.avg.2<- surv.avg.2 %>% 
  mutate(Condition.fc = (Condition.f * (length(Condition.f) - 1) + 0.5) / length(Condition.f))

# MODEL
## Betareg full model
## Residuals could not be improved by variable precision nor bias correction
bm1 <- betareg(Condition.fc ~ Treatment, data = surv.avg.2)

## Model selection fixed part
car::Anova(bm1) # Treatment has a significant effect (p=**)

```
## Model validation
```{r AVG survival: model validation}
##model validation of betareg model
#data is more normally distributed after square-transformation on Condition-data
mod <- bm1 # set betareg full model to be validated
op <- par(mfrow = c(2, 4), mar = c(5, 4, 1, 2)) # set layout of figure frame
plot(resid(mod, type = "pearson") ~ fitted(mod)) # fitted vs residuals
abline(0,0)
hist(resid(mod, type = "pearson"), main = "") # histogram residuals
qqnorm(resid(mod, type = "pearson")) # QQplot residuals
abline(0,0)
plot(surv.avg.2$Treatment, resid(mod, type = "pearson")) # residuals split over Treatment
abline(0,0)
plot(fitted(mod) ~ surv.avg.2$Condition.fc) # response data vs fitted
par(op)

```

## Post hoc and plots
```{r AVG survival: plots}

# Treatment plot (ignoring interactions)
# Get averages
surv.summary.T <- data_summary(surv.data.1, varname = "Condition", groupnames = c("Treatment"))

# Create ID and order dataframe by name of Finder
surv.summary.T <- tibble::rownames_to_column(surv.summary.T, "ID")
surv.summary.T <- surv.summary.T[order(as.character(surv.summary.T$Treatment)),]

# Post hoc comparison
hsd.surv.T <- emmeans(bm1, specs = pairwise ~ Treatment, adjust = "tukey") 
sig.letters.surv.T <- multcomp::cld(hsd.surv.T$emmeans, alpha = 0.05, Letters = letters, decreasing = T)

# Create ID and order dataframe by name (equals Finder of sgr.summary.SST dataframe)
sig.letters.surv.T <- sig.letters.surv.T[order(as.character(sig.letters.surv.T$Treatment)),]
sig.letters.surv.T <- sig.letters.surv.T %>% mutate(.group = str_squish(.group)) # remove white spaces

# Merge sig.letter dataframe into the summary dataframe
surv.summary.T <- cbind(surv.summary.T, siglet = sig.letters.surv.T$.group)
surv.summary.T <- surv.summary.T[order(as.numeric(surv.summary.T$ID)),]

# Plot Treatment bar graph + error bars + letters
SURV_Treatment<-ggplot(surv.summary.T, aes(x = Treatment, y = Condition))+
  geom_bar(stat = "identity", position = position_dodge())+ 
  labs(y = "Live tissue (%)", x = "Treatment")+
  geom_errorbar(aes(ymin=Condition-(1*se), ymax=Condition+(1*se)), width=.2,
                position=position_dodge(.9))+
  geom_text(data = surv.summary.T, aes(x=Treatment, y = Condition + se+4, label = siglet), vjust=0,
            position=position_dodge(.9))+
  theme_economist()+scale_colour_economist()+
  theme(
    axis.title.x = element_text(color="black", vjust=-2, size = 12, margin = margin(t = 10, r = 0, b = 0, l = 0)),
    axis.text.x = element_text(angle = 0, size=12, face = "bold", vjust=0.5),
    axis.title.y = element_text(color="black" , vjust=4, size = 12),
    axis.text.y=element_text(size=12, face = "bold", vjust=0.5))+
  scale_y_continuous(breaks= c(0, 25, 50, 75, 100), limits = c(0, 110), expand = c(0, 0))

SURV_Treatment
ggsave("Pd_Survival(AVG)_Treatment.pdf")

```

# Check packages used
```{r}

knitr::purl("Clean.Rmd")
list.functions.in.file("Clean.R")
unlink("Clean.R")

```

